---
title: "NYC flights + data wrangling"
webr:
  packages: ['tidyverse', 'scales', 'nycflights13']
filters:
  - webr
---

::: callout-important
Please reference the `webR-intro` programming exercise for information and instructions on how to interact with the programming exercise below.
:::

## Getting started

In this exercise, we will be exploring the `flights` data set, containing information about all flights that departed from NYC to destinations in the United States, Puerto Rico, and the American Virgin Island in 2013. A data key can be found below:

| Variable       | Description                                                            |
|----------------|------------------------------------------------------------------------|
| year           | year of departure                                                      |
| month          | month of departure                                                     |
| day            | day of departure                                                       |
| dep_time       | departure time                                                         |
| arr_time       | arrival time                                                           |
| sched_dep_time | scheduled departure time                                               |
| sched_arr_time | scheduled arrival time                                                 |
| dep_delay      | departure delays in minutes. Negative times represent early departures |
| arr_delay      | arrival delays in minutes. Negative times represent early arrivals     |
| carrier        | two letter carrier abbreviation                                        |
| tailnum        | plane tail number                                                      |
| orgin          | origin                                                                 |
| dest           | destination                                                            |
| air_time       | amount of time spent in the air in minutes                             |
| distance       | distance between airports in miles                                     |
| hour           | the hour of time scheduled departure                                   |
| minutes        | the minutes of time of scheduled departure                             |
| time_hour      | scheduled date and hour of the flight as a POSIXct date.            |

### Packages

We'll use: **tidyverse** for majority of the analysis and **scales** for pretty plot labels later on. These data we are using are in the **nycflights13** package. These are ready to use for you in this programming exercise!

```{r}
#| label: load-packages
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(scales)
library(nycflights13)
```

## Data wrangling 

Data wrangling is the process of transforming or reshaping data into a form that is more suitable for analysis. In this exercise, we will practice with a variety of functions used to wrangle data. 

## Exercise 1

Before we begin, use `glimpse` to report how many rows the `flights` data frame has. What does each row represent? 

```{webr-r}
# add code here
```

<br>
```{=html}
<textarea rows="4" cols="50">
</textarea>
```
<br><br>

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| echo: true
glimpse(flights)
```

Each observation represents a **flight**.
:::

{{< include _reflection_question.qmd >}}
