---
title: "Tidy Basketball"
toc: true
webr:
  packages: ['tidyverse', 'scales']
filters:
  - webr
---

::: callout-important
Please reference the `Meet the toolkit: Programming exercises` in `course 1` for information and instructions on how to interact with the programming exercise below.
:::

## Getting Started

```{r}
#| echo: false
#| include: false
library(tidyverse) # put in r file to r chunks have packages to use
library(scales)
library(ggridges)
library(gridExtra)
```


```{webr-r}
#| context: setup
download.file(
  "https://raw.githubusercontent.com/ElijahMeyer3/Coursera/main/data/trailblazer21.csv",
  "mariokart.csv"
)

options(readr.show_progress = FALSE)
```

```{webr-r}
#| echo: false
blazer <- read_csv("trailblazer21.csv")
```

```{r}
#| echo: false
#| warning: false
#| message: false
blazer <- read_csv("data/trailblazer21.csv") # put in R file to r chunks have data to use
```

The Portland Trailblazers are a National Basketball Association (NBA) sports team. In this mini analysis, we will work with a data set that has informatino of the points scored by 9 Portland Trailblazers players across the first 10 games of the 2021-2022 NBA season. If you would like to become more familiar with the game of basketball, please click [here](https://www.britannica.com/sports/basketball). If you would like more information on the Portland Trailblazers, please click [here](https://en.wikipedia.org/wiki/Portland_Trail_Blazers).